\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-08 chapters/patients.tex
Proofread

\Acp{bci} for communication assistive technology~\cite{Millan2010}
target individuals with \ac{sspi}~\cite{Peters2022}.
Visual \acp{bci}, which rely on the interpretation of visual stimuli by the user,
offer several advantages in this context.
They can work with non-invasive recording technology and can use rapid
stimulation.
This makes them well-suited for real-time tasks communication tasks.

Yet, there is a large comorbidity between \ac{sspi}, and eye motor impairment~\cite{FriedOken2020}.
Impairments such as nystagmus (uncontrolled eye movements), diplopia (double
vision), and ophthalmoplegia (eye paralysis) can significantly hinder
the ability to use visual BCIs. These impairments make it difficult for
\ac{bci} users to focus on or track visual stimuli accurately, reducing their
performance with BCIs that rely on visual cues~\cite{McCane2014,FriedOken2020,Pasqualotto2015}.
Unfortunately, it is again for this group that eye tracking solutions also
perform poorly, making them more reliant on potential developments in \ac{bci}
that do not rely on eye gaze.

Eye motor impairments are presumed to reduce performance in operating visual
oddball \ac{bci}s (see \autoref{sec:gaze-independence} for an overview), since users
cannot comfortably redirect their gaze at the desired target,
i.e., perform overt \ac{vsa}.
This is usually circumvented by designing gaze-independent~
\acp{bci}~\cite{Riccio2012}.
These interfaces either avoid visual stimulation, or exploit some form of
covert \ac{vsa}, where the gaze and \ac{vsa} do not coincide.

Several studies with visual oddball \acp{bci} show that performance drops when not fixating the intended
target~\cite{Brunner2010, Treder2010, RonAngevin2019}, necessitating
gaze-independent solutions.
These studies build on the assumption that \ac{bci} users with \ac{sspgi}
would feel comfortable operating an interface in pure covert \ac{vsa} with
central fixation.
One could argue that a \ac{bci} that is only verified to work when central
fixation is maintained, could also be considered gaze-dependent.
This does not account for the residual eye motor capabilities of most people
with \ac{sspi}, the (dis)comfort they experience while
performing gaze fixation and other confounding factors resulting from their eye
motility.

@@

\textcite{Lesenfants2014} tested a gaze-independent \ac{ssvep} \ac{bci} in six
participants with \ac{lis} yet only exceeded chance level accuracy in two.
More recently, \textcite{Peters2020} performed an trial with an
two participants with late-stage \ac{als} and visual impairment.
Their \ac{ssvep} interface was not optimized for gaze-independence, but the
system showed high accuracy, outperforming an eye tracking alternative.
It would be of interest to verify if such results can be verified with other
patient groups and visual oddball \acp{bci}.

\textcite{Orhan2012} and \textcite{Oken2014} tested the \ac{rsvp} speller with
individuals with \ac{lis}

@@

Our previous study, presented in chapter \autoref{sec:covert-align} also used the
visual Hex-o-Spell interface~\cite{VanDenKerchove2024}.
This work partially accounted for the idea that \ac{bci} users with \ac{sspgi} might
not fully rely on central gaze fixation and evaluates settings that are not
strictly dependent on this.
We showed gaze-independent performance can be improved
improved in healthy subjects using a suited decoding strategy.
that accounts for latency jitter in covert \ac{vsa} responses improves
Yet, there is a strong need for verification of these result in an applied
setting with people with \ac{sspi}.

%\todo{discuss \cite{FriedOken2020}}

Eventually, one of the end goals of this research line is to develop
gaze-independent \ac{bci} for peopl that are fully locked-in and have no option left than to use a \ac{bci}.
However, this group is very small and it is often a challenge to recruit them
into a study and perform experiments with them~\cite{Wolpaw2006}.
Individuals with less severe paralysis or in less progressed disease stages that struggle with
eye-tracking technology could also benefit from
solutions tailored to their specific situation.
Therefore, we aim to apply the concepts from earlier work an literature to
people with \ac{sspi} and various degrees motor impairment in a visuall oddball
\ac{bci}.
The objectives of this case study are as follows:
\begin{enumerate*}
  \item Explore capabilities and experienced comfort of individuals with
    \ac{sspgi}, when operating a visual \ac{bci},
  \item evaluate the performance of a gaze-independent visual \ac{bci} for this
    group,
  \item verify if this performance can be improved with a suitable decoding
    strategy.
\end{enumerate*}

\section{Materials \& methods}
\subsection{Recruitment}
Participants were recruited from across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France) and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Participants were recruited based on the following criteria.
To qualify for inclusion, participants must
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according the \ac{bci}	user selection criteria
    presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Participants were excluded if they
\begin{enumerate}
  \item have a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw},
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item have a predisposition to or have a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item have a severe loss in vision or hearing, that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the out-
        come.\label{item:patients/exclusion/cognitive}
  \item be able to understand the experiment instructions and cooperate,
  \item have any other limitations preventing them from performing the given task.
\end{enumerate}

In total, 11 individuals were contacted, of which one person with \ac{ms} was excluded based on
criterion~\ref{item:patients/exclusion/vision}, one person recovering from
\ac{tbi} on
both~\ref{item:patients/exclusion/epilepsy}
and~\ref{item:patients/exclusion/cognitive}, and one person recovering from
stroke based
on~\ref{item:patients/exclusion/medical}
One further person recovering from a stroke was excluded due to technical
difficulties during the experimental session.
Vision was assessed using a LogMAR chart~\cite{Bailey1976}.

Ultimately, 7 participants were retained.
Of these, one participant was diagnosed from bulbar-onset \ac{als}.
\ac{als} is a neurodegenerative disease affecting the motor neurons, leading to
progressive loss of motor function.
This initially results in general weakness and loss of muscle tone, but
eventually results in full body paralysis.
Although speech and especially eye movements are usually preserved until the
later stages of the disease progression, the bulbar-onset variant is
characterized by an early loss of speech and an increased involvement of eye
motor symptoms~\cite{Guo2022}.
Furthermore, one of the goals of \ac{bci} is to support individuals with \ac{als}
whose life span has been extended with life support with and assistive technology to
ensure quality of life.
In these very progressed stages of \ac{als}, eye movement will eventually also
be affected~\cite{Hayashi1991}.

Three other participants were diagnosed with \ac{fa}, a neurodegenerative
disease affecting the
spinal cord, peripheral nervous system and cerebellum.
\ac{fa} results in an impairing loss of muscle coordination.
In progressed stages, the disease can present with nystagmus, saccadic
intrusions and gaze dysmetria~\cite{Cook2017}.

@@

\newcommand{\fnkuebler}{%
  \footnote{
``With minor degree of impairment,
we refer to patients who had only slightly impaired limb movement
and normal speech. Under the category moderate impairment, we
summarized those patients with restricted limb movement (wheelchair-bound) and
unaffected speech or intact limb movement without speech (such as the bulbar form of ALS which first affects speech
and swallowing). Patients who were almost tetraplegic with restricted speech were considered major impaired. Categories four
and five were the LIS and the CLIS, respectively."~\cite{Kuebler2008}
  }
}
\begin{table}[t]
  \centering
  \makebox[\textwidth][c]{%
  \footnotesize
  \input{chapters/patients_table.tex}
  }
  \caption[Included participants with their diagnosis and
  capabilities.]{Included participants with their diagnosis and
  capabilities.
  Trach.: underwent a tracheostomy, Cls. W: classification according
  to~\textcite{Wolpaw2006}, Cls. KB: classification according
  to~\textcite{Kuebler2008}\fnkuebler.
  }
  \label{tab:patients/patients}
\end{table}
\todo{include years since diagnosis}

@@

Self-reported eye motor and visual abnormalities were recorded according to the
relevant visual \ac{bci} skills presented by~\textcite{FriedOken2020}.
These include visual acuity, visual fixation, eyelid function, ocular motility,
binocular vision and field of vision.
Additionally, participants were asked about eye tremors (nystagmus or other) and
other involuntary eye movements.

@@

Table~\ref{tab:patients/eye} details their eye motor impairments and vision of
included participants.
All participants reported some degree of fatigue or discomfort when fixating.
Participant PA1 had the mildest impairment, only reporting fatigue when fixating
for prolonged times.
The \ac{fa} participants were mostly affected by eye tremors and impaired pursuit.
PB2 suffered from especially severe horizontal oscillating involuntary eye
movements.
Eye motor function of participants PC2, PC3, and PC3 were most severly affected.
Participant PC2 was only able to look up and down and had a deviation in the
left eye causing diplopia, but this was corrected by a prism glass.
Participant PC3 only retained partial motility of the right eye, while the left eye was permanently closed.
Participant PC4 had one deviated eye with a corneal abcess affecting the motility
and vision in the right eye, and suffered reduced motility in the left.

@@

Finally, we also record gaze position throughout the experimental session, to
register the participant's gaze relative to the stimulated \ac{bci} targets.

@@

The \ac{bci} stimulation procedure was based on the
Hex-o-Spell~\cite{Treder2010} implementation presented
by~\textcite{VanDenKerchove2024}.
Similar to this study, the task consists of counting the flashes of a cued
target among 6 round, flashing targets laid out in a hexagonal pattern in the
field of view of the user.
We refer to section~\ref{sec:covert-align/stimulation} for implementation details.

3 different \ac{vsa} settings were explored
In the overt \ac{vsa} setting, the participant was instructed to fixate the cued target, or
try to the maximum extent of their visual skill, even if experiencing slight
discomfort.
In the covert \ac{vsa} setting, the participant was instructed to fixate the center of the
screen, also to the extent of their ability.
An additional \emph{free \ac{vsa}} setting was introduced.
Here, the participant was instructed to perform the task as they deemed most
comfortable.
This allows us to investigate the user's natural way of operating the \ac{bci}
given their individual set of visual skills.
If the participant was not fully paralyzed, they were instructed not to move the head.
The cued \emph{split attention} setting proposed
by~\textcite{VanDenKerchove2024} is not studied here, since we are interested
in natural \ac{vsa} operation settings for gaze-impaired individuals.

To make the interface suitable for use by individuals with
\ac{sspi}~\cite{FriedOken2020}, the
number of blocks was decreased to 6 per \ac{vsa} setting.
\Ac{isi} was increased to $200\pm50$ms to decrease task difficulty.
The experiment also started with a training block in each condition, where the
participant was instructed with feedback on their performance to make sure they
understood and were able to perform the task.

@@

During the recording session, participants were positioned in their wheelchair in front of a table.
Stimuli were presented on a Acer Predator Helios laptop with 18" screen (Acer,
Inc., Taiwan) placed at a 60cm distance.
A Cedrus StimTracker (Cedrus Corp., CA, USA) ensured synchronization of stimuli with the
recorded \ac{eeg}.
Eye tracking throughout was performed using the Tobii X2-30 Compact (Tobii
Technology AB, Sweden) portable eye tracker placed at the bottom of the laptop screen.

\Ac{eeg} was recorded at 1000Hz using the Neuroscan Neuvo  portable amplifier (Compumedics Neuroscan,
Australia) connected to a second laptop for registration
The \ac{eeg} headset used 18 active AgCl electrodes (EASYCAP GmbH, Germany) placed on a cap
according to the international 10-20 layout.
Using electrolyte gel, electrode impedances were reduced below 10k$\Omega$
Additionally, the \ac{eog} was recorded.

The \ac{eeg} was band-pass filtered between 0.5 and 16Hz.
Bad channels were rejected using on the RANSAC algorithm~\cite{Fischler1981}
and visual inspection.
Next, the \ac{eeg} was rereferenced to the average of mastoid electrodes TP9
and TP10, and \ac{ica} was performed to reject artifactual components based on
correlation with the \ac{eog} or by visual inspection.
Epochs are cut from -0.1 to 0.9s relative to stimulus onse, no baseline
correction was performed.

Eye tracking data was cleaned by fusing left and right gaze into one channel
for the horizontal and vertical gaze position.
If both were present for a given sample, the fused channel was the mean of both
values.
If at a given sample either the left or the right eye was not detected for a
given channel, the value of the other one was adopted.
If both were missing, the gaze position remains unset at that time point and no
interpolation is performed.

@@

We evaluate the recorded data using the \ac{wcble}~\cite{VanDenKerchove2024}
and \ac{tlda}~\cite{Sosulski2022}.
classifiers, and the Riemannian XDAWNCov+TS+LDA~\cite{Cecotti2017}.
For \ac{wcble}, a region of interest from 0ms to 800ms relative to stimulus
onset was used while the epoch was cropped to -100ms to 900ms, for the other
decoders the epoch was cropped between 0ms and 800ms which resulted in maximal
performance.
Decoding scores were obtained using 6-fold cross-validation where folds corresponded to
stimulation blocks.

refer to~\autoref{sec:covert-align/methods/decoders}.

@@

First, we aimed to shed more light on the actual capabilities of individuals
with \ac{sspgi} with respect to performing overt \ac{vsa} and central gaze
fixation, as well as investigate how relevant these two settings are when the
gaze is not cued.
Figure~\ref{fig:patients/gaze} maps gaze position relative to the stimuli
across conditions.
These results should be interpreted with care, since the eye tracker to some
degree relies on functioning eye motility.
The participant's position
relative due the eye tracker might have shifted throughout the experimental
session despite our best efforts, e.g.\ because they needed aspiration of their
tracheostomy.

@@

PA1 had relatively intact gaze control and was able to correctly perform the
cued overt and covert settings.
When gaze was uncued he fixated the cued target.
This was also mostly the case for PB1, although eye tracking revealed that he
chose not to perform central gaze fixation when cued in at least one of the
stimulation blocks. We were unable to record his gaze near the bottom left
stimulus position, either due to eye tracker failure or because the participant
was not comfortable fixating this position.
Eye tracker calibration did not succeed for subject PB4, but given
transformations of gaze positions to the stimulus space, thy

PB2 was able to perform overt \ac{vsa} and central fixation to some extent,
yet eye tracking, but gaze position distributions shows a larger spread than
for PA1 and PB1.
The distribution in free \ac{vsa}, however, showed that she prefered to attend
stimuli covertly when the gaze was uncued.
This was confirmed by the participant.

The overt and central gaze fixation settings were also not properly adapted to
particpant PC4.
In free \ac{vsa} eye tracker results show that his gaze was usually near the
bottom two targets, indicating some degree of covert or split \ac{vsa}.

It was technically impossible to register gaze position with the Tobii X2-30
Compact for participants PC2 and PC3 since they both had one eye that was
occluded respectively by the prism glass and the eyelid.
Both participants reported they could not fixate some of the
stimuli.

@@

Figure~\ref{fig:patients/decode} shows single-trial
\ac{rocauc} in the evaluated \ac{vsa} settings fort the different decoders.

\begin{figure}[t]
  \input{figures/patients/fig_decode.pgf}
  \caption[%
    Decoding performance in different \acs{vsa} settings.
  ]{%
    Decoding performance in different \ac{vsa} settings reported as
    single-trial \ac{rocauc}.
    Free \ac{vsa} is generally on par with performance in the overt \ac{vsa}
    settign.
    Performance in the covert \ac{vsa} setting with central gaze fixation is lower, but can
    be improved with the \ac{wcble} decoder.
    95\% confidence intervals were calculated using 1000 bootstrapping
    repetitions.
  }
  \label{fig:patients/decode}
\end{figure}
\todo{check if percentages or 0-1 accuracy and roc-auc everywhere}

In the overt \ac{vsa} setting, evaluated decoders performed about equal on average
(\ac{wcble} 75.58\%, XDAWNCov+TS+LDA 74.24\%, \ac{tlda} 75.99\%).
In the covert \ac{vsa} setting with cued central gaze fixation, performance deteriorates
but \ac{wcble} can significantly improve performance over its base classifier
\ac{tlda} in this condition
(\ac{wcble} 62.49\%, XDAWNCov+TS+LDA 59.42\%, \ac{tlda} 59.05\%).
Decoding performance for this task was at chance level for participants PB4 and
PC3.

However, \Ac{wcble} did not improve \ac{tlda} in free \ac{vsa} free \ac{vsa}, but
XDAWNCov+TS+LDA performance was slightly lower here (although not
significantly).
(\ac{wcble} 74.15\%, XDAWNCov+TS+LDA 71.88\%, \ac{tlda} 74.27\%).
More interestingly, we notice that performances of the decoders in free
\ac{vsa} are close to those in the overt \ac{vsa}.
A substantial decrease in performance from the overt setting to the free
setting is present for
subjects PC3 and PC4.
For PB2, who also relied on covert \ac{vsa} during the uncued free \ac{vsa}
setting, there was no decrease in performance.

\subsection{Cross-condition calibration}
\label{sec:patients/outcomes/cross}
As an alternative approach to picking the most suitable decoder, we chose
\ac{tlda} as a given decoder and verified whether performance can be improved
if \ac{bci} users with gaze impairment perform the calibration session relying
maximally on their residual gaze control.

>>>>

\Acp{bci} for communication assistive technology~\cite{Millan2010}
target individuals with \acf{sspi}~\cite{Peters2022}.
Visual \acp{bci}, which rely on the interpretation of visual stimuli by the user,
offer several advantages in this context.
They can work with non-invasive recording technology and can use rapid
stimulation.
This makes them well-suited for real-time communication tasks.

Yet, there is a large comorbidity between \ac{sspi} and eye motor impairment~\cite{FriedOken2020}.
Impairments such as nystagmus (uncontrolled eye movements), diplopia (double
vision), and ophthalmoplegia (eye paralysis) can significantly hinder
the ability to use visual BCIs. These impairments make it difficult for
\ac{bci} users to focus on or track visual stimuli accurately, reducing their
performance with BCIs that rely on visual cues~\cite{McCane2014,FriedOken2020,Pasqualotto2015}.
Unfortunately, it is again for this group that eye tracking solutions also
perform poorly, making them more reliant on potential developments in \ac{bci}
that do not rely on eye gaze.

Eye motor impairments are presumed to reduce performance in operating visual
oddball \ac{bci}s (see \cref{sec:gaze-independence} for an overview), since users
cannot comfortably redirect their gaze at the desired target,
i.e., perform overt \ac{vsa}.
This is usually circumvented by designing gaze-independent~
\acp{bci}~\cite{Riccio2012}.
These interfaces either avoid visual stimulation or exploit some form of
covert \ac{vsa}, where the gaze and \ac{vsa} do not coincide.

Several studies with visual oddball \acp{bci} show that performance drops when not fixating the intended
target~\cite{Brunner2010, Treder2010, RonAngevin2019}, necessitating
gaze-independent solutions.
These studies build on the assumption that \ac{bci} users with \ac{sspgi}
would feel comfortable operating an interface in pure covert \ac{vsa} with
central fixation.
One could argue that a \ac{bci} that is only verified to work when central
fixation is maintained could also be considered gaze-dependent.
This does not account for the residual eye motor capabilities of most people
with \ac{sspi}, the (dis)comfort they experience while
performing gaze fixation and other confounding factors resulting from their eye
motility.

@@

\textcite{Lesenfants2014} tested a gaze-independent \ac{ssvep} \ac{bci} in six
participants with \ac{lis} yet only exceeded chance level accuracy in two.
More recently, \textcite{Peters2020} performed a trial with two
participants with late-stage \ac{als} and visual impairment.
Their \ac{ssvep} interface was not optimized for gaze-independence, but the
system showed high accuracy, outperforming an eye tracking alternative.
It would be of interest to verify if such results can be replicated with other
patient groups and visual oddball \acp{bci}.

\textcite{Orhan2012} and \textcite{Oken2014} tested the \ac{rsvp} speller with
individuals with \ac{lis}.

@@

Our previous study, presented in \cref{sec:covert-align} also used the
visual Hex-o-Spell interface~\cite{VanDenKerchove2024}.
This work partially accounted for the idea that \ac{bci} users with \ac{sspgi} might
not fully rely on central gaze fixation and evaluated settings that are not
strictly dependent on this.
We showed gaze-independent performance can be improved in healthy subjects
using a suited decoding strategy.
That accounts for latency jitter in covert \ac{vsa} responses.
Yet, there is a strong need for verification of these results in an applied
setting with people with \ac{sspi}.

Eventually, one of the end goals of this research line is to develop
gaze-independent \ac{bci} for people that are fully locked-in and have no option left than to use a \ac{bci}.
However, this group is very small and it is often a challenge to recruit them
into a study and perform experiments with them~\cite{Wolpaw2006}.
Individuals with less severe paralysis or in less progressed disease stages that struggle with
eye-tracking technology could also benefit from
solutions tailored to their specific situation.
Therefore, we aim to apply the concepts from earlier work and literature to
people with \ac{sspi} and various degrees of motor impairment in a visual oddball
\ac{bci}.
The objectives of this case study are as follows:
\begin{enumerate*}
  \item Explore capabilities and experienced comfort of individuals with
    \ac{sspgi}, when operating a visual \ac{bci},
  \item evaluate the performance of a gaze-independent visual \ac{bci} for this
    group,
  \item verify if this performance can be improved with a suitable decoding
    strategy.
\end{enumerate*}

\section{Materials \& methods}
\subsection{Recruitment}
Participants were recruited from across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France), and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Participants were recruited based on the following criteria.
To qualify for inclusion, participants must:
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according to the \ac{bci}	user selection criteria
    presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Participants were excluded if they:
\begin{enumerate}
  \item have a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw},
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item have a predisposition to or have a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item have a severe loss in vision or hearing that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the outcome.\label{item:patients/exclusion/cognitive}
  \item be unable to understand the experiment instructions and cooperate,
  \item have any other limitations preventing them from performing the given task.
\end{enumerate}

In total, 11 individuals were contacted, of which one person with \ac{ms} was excluded based on
criterion~\ref{item:patients/exclusion/vision}, one person recovering from
\ac{tbi} on
both~\ref{item:patients/exclusion/epilepsy}
and~\ref{item:patients/exclusion/cognitive}, and one person recovering from
stroke based
on~\ref{item:patients/exclusion/medical}.
One further person recovering from a stroke was excluded due to technical
difficulties during the experimental session.
Vision was assessed using a LogMAR chart~\cite{Bailey1976}.

Ultimately, 7 participants were retained.
Of these, one participant was diagnosed with bulbar-onset \ac{als}.
\ac{als} is a neurodegenerative disease affecting the motor neurons, leading to
progressive loss of motor function.
This initially results in general weakness and loss of muscle tone, but
eventually leads to full body paralysis.
Although speech and especially eye movements are usually preserved until the
later stages of the disease progression, the bulbar-onset variant is
characterized by an early loss of speech and an increased involvement of eye
motor symptoms~\cite{Guo2022}.
Furthermore, one of the goals of \ac{bci} is to support individuals with \ac{als}
whose life span has been extended with life support and assistive technology to
ensure quality of life.
In these very progressed stages of \ac{als}, eye movement will eventually also
be affected~\cite{Hayashi1991}.

Three other participants were diagnosed with \ac{fa}, a neurodegenerative
disease affecting the
spinal cord, peripheral nervous system, and cerebellum.
\ac{fa} results in an impairing loss of muscle coordination.
In progressed stages, the disease can present with nystagmus, saccadic
intrusions, and gaze dysmetria~\cite{Cook2017}.

@@

\newcommand{\fnkuebler}{%
  \footnote{
``With minor degree of impairment,
we refer to patients who had only slightly impaired limb movement
and normal speech. Under the category moderate impairment, we
summarized those patients with restricted limb movement (wheelchair-bound) and
unaffected speech or intact limb movement without speech (such as the bulbar form of ALS which first affects speech
and swallowing). Patients who were almost tetraplegic with restricted speech were considered majorly impaired. Categories four
and five were the LIS and the CLIS, respectively."~\cite{Kuebler2008}
  }
}
\begin{table}[t]
  \centering
  \makebox[\textwidth][c]{%
  \footnotesize
  \input{chapters/patients_table.tex}
  }
  \caption[Included participants with their diagnosis and
  capabilities.]{Included participants with their diagnosis and
  capabilities.
  Trach.: underwent a tracheostomy, Cls. W: classification according
  to~\textcite{Wolpaw2006}, Cls. KB: classification according
  to~\textcite{Kuebler2008}\fnkuebler.
  }
  \label{tab:patients/patients}
\end{table}
%\todo{include years since diagnosis}

@@

Self-reported eye motor and visual abnormalities were recorded according to the
relevant visual \ac{bci} skills presented by~\textcite{FriedOken2020}.
These include visual acuity, visual fixation, eyelid function, ocular motility,
binocular vision, and field of vision.
Additionally, participants were asked about eye tremors (nystagmus or other) and
other involuntary eye movements.

@@

Table~\ref{tab:patients/eye} details the eye motor impairments and vision of
the included participants.
All participants reported some degree of fatigue or discomfort when fixating.
Participant PA1 had the mildest impairment, only reporting fatigue when fixating
for prolonged times.
The \ac{fa} participants were mostly affected by eye tremors and impaired pursuit.
PB2 suffered from especially severe horizontal oscillating involuntary eye
movements.
Eye motor function of participants PC2, PC3, and PC4 was most severely affected.
Participant PC2 was only able to look up and down and had a deviation in the
left eye causing diplopia, but this was corrected by a prism glass.
Participant PC3 only retained partial motility of the right eye, while the left eye was permanently closed.
Participant PC4 had one deviated eye with a corneal abscess affecting the motility
and vision in the right eye, and suffered reduced motility in the left.

@@

Finally, we also recorded gaze position throughout the experimental session to
register the participant's gaze relative to the stimulated \ac{bci} targets.

@@

The \ac{bci} stimulation procedure was based on the
Hex-o-Spell~\cite{Treder2010} implementation presented
by~\textcite{VanDenKerchove2024}.
Similar to this study, the task consists of counting the flashes of a cued
target among 6 round, flashing targets laid out in a hexagonal pattern in the
field of view of the user.
We refer to \cref{sec:covert-align/stimulation} for implementation details.

Three different \ac{vsa} settings were explored.
In the overt \ac{vsa} setting, the participant was instructed to fixate on the cued target or
try to the maximum extent of their visual skill, even if experiencing slight
discomfort.
In the covert \ac{vsa} setting, the participant was instructed to fixate on the center of the
screen, to the extent of their ability.
An additional \emph{free \ac{vsa}} setting was introduced.
Here, the participant was instructed to perform the task as they deemed most
comfortable.
This allowed us to investigate the user's natural way of operating the \ac{bci}
given their individual set of visual skills.
If the participant was not fully paralyzed, they were instructed not to move their head.
The cued \emph{split attention} setting proposed
by~\textcite{VanDenKerchove2024} was not studied here, as we were interested
in natural \ac{vsa} operation settings for gaze-impaired individuals.

To make the interface suitable for use by individuals with
\ac{sspi}~\cite{FriedOken2020}, the
number of blocks was decreased to 6 per \ac{vsa} setting.
\Ac{isi} was increased to $200\pm50$ms to decrease task difficulty.
The experiment also started with a training block in each condition, where the
participant was instructed with feedback on their performance to ensure they
understood and were able to perform the task.

@@

During the recording session, participants were positioned in their wheelchair in front of a table.
Stimuli were presented on an Acer Predator Helios laptop with an 18" screen (Acer,
Inc., Taiwan) placed at a 60 cm distance.
A Cedrus StimTracker (Cedrus Corp., CA, USA) ensured synchronization of stimuli with the
recorded \ac{eeg}.
Eye tracking was performed throughout using the Tobii X2-30 Compact (Tobii
Technology AB, Sweden) portable eye tracker placed at the bottom of the laptop screen.

\Ac{eeg} was recorded at 1000 Hz using the Neuroscan Neuvo portable amplifier (Compumedics Neuroscan,
Australia) connected to a second laptop for registration.
The \ac{eeg} headset used 18 active AgCl electrodes (EASYCAP GmbH, Germany) placed on a cap
according to the international 10-20 layout.
Using electrolyte gel, electrode impedances were reduced below 10k$\Omega$.
Additionally, the \ac{eog} was recorded.

The \ac{eeg} was band-pass filtered between 0.5 and 16 Hz.
Bad channels were rejected using the RANSAC algorithm~\cite{Fischler1981}
and visual inspection.
Next, the \ac{eeg} was re-referenced to the average of mastoid electrodes TP9
and TP10, and \ac{ica} was performed to reject artifactual components based on
correlation with the \ac{eog} or by visual inspection.
Epochs were cut from -0.1 to 0.9s relative to stimulus onset, and no baseline
correction was performed.

Eye tracking data was cleaned by fusing left and right gaze into one channel
for the horizontal and vertical gaze position.
If both were present for a given sample, the fused channel was the mean of both
values.
If at a given sample either the left or the right eye was not detected for a
given channel, the value of the other one was adopted.
If both were missing, the gaze position remained unset at that time point, and no
interpolation was performed.

@@

We evaluated the recorded data using the \ac{wcble}~\cite{VanDenKerchove2024}
and \ac{tlda}~\cite{Sosulski2022}
classifiers, as well as the Riemannian XDAWNCov+TS+LDA~\cite{Cecotti2017}.
For \ac{wcble}, a region of interest from 0 ms to 800 ms relative to stimulus
onset was used while the epoch was cropped to -100 ms to 900 ms. For the other
decoders, the epoch was cropped between 0 ms and 800 ms, which resulted in maximal
performance.
Decoding scores were obtained using 6-fold cross-validation where folds corresponded to
stimulation blocks.

@@

First, we aimed to shed more light on the actual capabilities of individuals
with \ac{sspgi} regarding performing overt \ac{vsa} and central gaze
fixation, as well as to investigate how relevant these two settings are when the
gaze is not cued.
\Cref{fig:patients/gaze} maps gaze position relative to the stimuli
across conditions.
These results should be interpreted with care, as the eye tracker to some
degree relies on functioning eye motility.
The participant's position relative to the eye tracker might have shifted
throughout the experimental session despite our best efforts, e.g., because they needed aspiration of their
tracheostomy.

@@

PA1 had relatively intact gaze control and was able to correctly perform the
cued overt and covert settings.
When gaze was uncued, he fixated on the cued target.
This was also mostly the case for PB1, although eye tracking revealed that he
chose not to perform central gaze fixation when cued in at least one of the
stimulation blocks. We were unable to record his gaze near the bottom-left
stimulus position, either due to eye tracker failure or because the participant
was not comfortable fixating on this position.
Eye tracker calibration did not succeed for subject PB4, but given
transformations of gaze positions to the stimulus space, they were assumed to be overtly performing the task.

PB2 was able to perform overt \ac{vsa} and central fixation to some extent,
yet eye tracking shows a larger spread in gaze position compared to
PA1 and PB1.
In the free \ac{vsa} condition, however, she preferred to attend
stimuli covertly when the gaze was uncued.
This was confirmed by the participant.

The overt and central gaze fixation settings were also not properly adapted to
participant PC4.
In the free \ac{vsa} condition, eye tracker results show that his gaze was usually near the
bottom two targets, indicating some degree of covert or split \ac{vsa}.

It was technically impossible to register gaze position with the Tobii X2-30
Compact for participants PC2 and PC3 since they both had one eye that was
occluded respectively by the prism glass and the eyelid.
Both participants reported they could not fixate on some of the
stimuli.

@@

\Cref{fig:patients/decode} shows single-trial
\ac{rocauc} in the evaluated \ac{vsa} settings for the different decoders.

\begin{figure}[t]
  \input{figures/patients/fig_decode.pgf}
  \caption[%
    Decoding performance in different \acs{vsa} settings.
  ]{%
    Decoding performance in different \ac{vsa} settings reported as
    single-trial \ac{rocauc}.
    Free \ac{vsa} is generally on par with performance in the overt \ac{vsa}
    setting.
    Performance in the covert \ac{vsa} setting with central gaze fixation is lower, but can
    be improved with the \ac{wcble} decoder.
    95\% confidence intervals were calculated using 1000 bootstrapping
    repetitions.
  }
  \label{fig:patients/decode}
\end{figure}

In the overt \ac{vsa} setting, the evaluated decoders performed similarly on average
(\ac{wcble} 75.58\%, XDAWNCov+TS+LDA 74.24\%, \ac{tlda} 75.99\%).
In the covert \ac{vsa} setting with cued central gaze fixation, performance deteriorated,
but \ac{wcble} significantly improved performance over the base classifier
\ac{tlda} in this condition
(\ac{wcble} 62.49\%, XDAWNCov+TS+LDA 59.42\%, \ac{tlda} 59.05\%).
Decoding performance for this task was at chance level for participants PB4 and
PC3.

However, \ac{wcble} did not improve \ac{tlda} performance in the free \ac{vsa} setting, but
XDAWNCov+TS+LDA performance was slightly lower here (though not
significantly).
(\ac{wcble} 74.15\%, XDAWNCov+TS+LDA 71.88\%, \ac{tlda} 74.27\%).
More interestingly, we noticed that performances of the decoders in free
\ac{vsa} were close to those in the overt \ac{vsa}.
A substantial decrease in performance from the overt setting to the free
setting was observed for
subjects PC3 and PC4.
For PB2, who also relied on covert \ac{vsa} during the uncued free \ac{vsa}
setting, there was no decrease in performance.

\subsection{Cross-condition calibration}
\label{sec:patients/outcomes/cross}

As an alternative approach to selecting the most suitable decoder, we used
\ac{tlda} as the base decoder and verified whether performance could be improved
if \ac{bci} users with gaze impairment performed the calibration session relying
maximally on their residual gaze control.

\Cref{fig:patients/cross} shows that, on average, covert \ac{vsa} decoding
improved when training with overt \ac{vsa}.
This was especially true for participants PA1, PB2, and PC3.
Note that, according to eye tracking data, participants PB1, PB4, and PC4 did not
always perform cued central gaze fixation in the covert \ac{vsa} setting,
which might have affected the results.

\begin{figure}[t]
  \input{figures/patients/fig_cross.pgf}
  \caption[Cross-setting calibration and decoding performance.]{
    Decoding performance when calibrating the \ac{tlda} decoder in a given \ac{vsa}
    setting, and evaluating it in another, reported as
    single-trial \ac{rocauc}.
    For participant PA1, free \ac{vsa} performance improved when
    calibrated with overt gaze fixation.
    For participants PA1, PB2, and PC3, performance in the covert \ac{vsa} setting with central gaze fixation
    improved when calibrating with overt gaze fixation.
  }
  \label{fig:patients/cross}
\end{figure}
